Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-12-20 threshold/draft.tex
Added raw tex/image files for jss submission
old R files/example moving to package

\begin{figure}[!ht]
\begin{center}
<<eval=TRUE, echo=FALSE, fig=TRUE, height=5, width=7>>=
library(biclust)
data(EisenYeast)
A = t(EisenYeast)
A = sweep(A, 2, apply(A,2,mean), FUN=`-`)
A = sweep(A, 2, sqrt(apply(A,2,crossprod)), FUN=`/`)
s = svd(A)
i = order(s$v[,1])
plot(s$v[i,1], ylab=expression(P * v[1]), xlab="Sorted index", pch='.')
t=0.99
abline(h=c(0, sqrt(2*(1-t))/s$d[1]), col=2, lwd=1)
@
\caption{
Example plot of the ordered entries of $v_1$ for the EisenYeast example from
Section \ref{examples}. The red lines illustrate the interval corresponding
threshold $t=0.99$ placed at an arbitrary location on the vertical axis. Points
farther apart than this interval correspond to column vectors that do not meet
the correlation threshold.
\label{fig1}
}
\end{center}
\end{figure}

@@

<<eval=FALSE,echo=TRUE>>=
devtools::install_github("bwlewis/tcor")
@

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
<<eval=TRUE,echo=TRUE>>=
library(tcor)
library(doMC)     # local parallel computation
library(biclust)  # for the BicatYeast data
data(EisenYeast)  # in gene by sample orientation
A = t(EisenYeast) # to correlate genes to genes
threshold = 0.95

# Register a four-core multicore parallel backend
registerDoMC(4)

# Compute using tcor():
m1  = sum(gc()[,2])  # Memory in megabytes currently in use
t1  = proc.time()
tx  = tcor(A, t=threshold, p=10)
t1  = (proc.time() - t1)[3] # Time in seconds
m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test

# Fast brute force full correlation:
m2  = sum(gc()[,2])
t2  = proc.time()
mu = colMeans(A)
s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
cs = scale(A, center=mu, scale=s)  # scale and center the data
cx = crossprod(cs)                 # full correlation matrix
cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
pairs = which(cx >= threshold, arr.ind=TRUE)
t2  = (proc.time() - t2)[3]
m2  = sum(gc()[,6]) - m2
@
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

<<echo=FALSE, results=tex>>=
require(xtable)
ans = data.frame(Method=c("tcor","brute force"), `Wall clock time (s)`=c(t1, t2), `Peak excess memory use (MB)`=c(m1, m2), check.names=FALSE)
print(
    xtable(ans,
        caption = "Gene expression results, threshold=0.95", label = "EisenYeast", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE)
@

@@

<<echo=FALSE, results=tex>>=
# Test system values:
t1=10309
t3=482
t4=NA
m1=389.1
m3=994.3
m4=42282.4

if(Sys.getenv("RUN_EXAMPLES")=="TRUE")
{
  library(doMC)
  system("wget http://gdac.broadinstitute.org/runs/stddata__2015_08_21/data/ACC/20150821/gdac.broadinstitute.org_ACC.Merge_methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.Level_3.2015082100.0.0.tar.gz")
  cmd = "cat ACC.methylation__humanmethylation450__jhu_usc_edu__Level_3__within_bioassay_data_set_function__data.data.txt  | tail -n +%s | cut -d '       ' -f 2,6,10,14,18,22,26,30,34,38,42,46,50,54,58,62,66,70,74,78,82,86,90,94,98,102,106,110,114,118,122,126,130,134,138,142,146,150,154,158,162,166,170,174,178,182,186,190,194,198,202,206,210,214,218,222,226,230,234,238,242,246,250,254,258,262,266,270,274,278,282,286,290,294,298,302,306,310,314,318"
  N = 485579  # total number of lines
  h = 2       # total header lines
  # Read the first header line with sample IDs:
  f = pipe(cmd, open="r")
  id = as.vector(read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=1))
  close(f)
  # Read the rest of the file in parallel. I prefer doMC/foreach over mclapply for this because foreach can combine results incrementally.
  registerDoMC(4)  # four cores on our test machine
  block = floor((N - h)/cores)
  acc = foreach(j=1:cores, .combine=rbind) %dopar%
  {
    skip = block * (j - 1) + h + 1
    nrows = ifelse(j == cores, -1, block)
    f = pipe(sprintf(cmd, skip), open="r")
    on.exit(close(f))
    read.table(f, sep="\t", stringsAsFactors=FALSE, header=FALSE, nrows=nrows)
  }
  acc = t(acc)
  # OK, the example data are parsed and in acc. proceed as in the smaller example.
  threshold = 0.99
  # Compute using tcor():
  m1  = sum(gc()[,2])  # Memory in megabytes currently in use
  t1  = proc.time()
  tx  = tcor(acc, t=threshold, p=10)
  t1  = (proc.time() - t1)[3] # Time in seconds
  m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test

  # The cluster example proceeds similarly, but registers a doRedis backend
  # instead of the doMC backend before invoking tcor...
} # end if run examples
require(xtable)
ans = data.frame(Method=c("tcor (4 core PC)", "tcor (64 core cluster)", ""), `Wall clock time (s)`=c(t1, t3, t4), `Memory use (MB)`=c(m1, m3, m4), " "=c("","(local)","(remote)"), check.names=FALSE)
print(
    xtable(ans,
        caption = "TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).", label = "TCGA3", caption.placement = "top")
  ,size="small", table.placement="ht", include.rownames=FALSE, NA.string="")
@
From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.

>>>>

\begin{figure}[!ht]
\begin{center}
\includegraphics{draft-001}
\caption{
Example plot of the ordered entries of $v_1$ for the EisenYeast example from
Section \ref{examples}. The red lines illustrate the interval corresponding
threshold $t=0.99$ placed at an arbitrary location on the vertical axis. Points
farther apart than this interval correspond to column vectors that do not meet
the correlation threshold.
\label{fig1}
}
\end{center}
\end{figure}

@@

\begin{Schunk}
\begin{Sinput}
> devtools::install_github("bwlewis/tcor")
\end{Sinput}
\end{Schunk}

@@

Anyone can quickly run the example to experiment with the algorithm and
compare its output with brute force results. The example uses the {\tt
EisenYeast} data from the {\tt biclust} \proglang{R} package \citep{biclust}.
The data consist of 80 experiments involving the
Saccharomyces Cerevisiae yeast measuring gene expression data across 6,221
Affymetrix probes, represented as an $80\times 6221$ real-valued matrix.  The
example finds all pairs of columns (gene expression measurements) that meet or exceed a
Pearson correlation value of 0.95.  We use a projected dimension $p=10$
in {\tt tcor()}.
\begin{small}
\begin{Schunk}
\begin{Sinput}
> library(tcor)
> library(doMC)     # local parallel computation
> library(biclust)  # for the BicatYeast data
> data(EisenYeast)  # in gene by sample orientation
> A = t(EisenYeast) # to correlate genes to genes
> threshold = 0.95
> # Register a four-core multicore parallel backend
> registerDoMC(4)
> # Compute using tcor():
> m1  = sum(gc()[,2])  # Memory in megabytes currently in use
> t1  = proc.time()
> tx  = tcor(A, t=threshold, p=10)
> t1  = (proc.time() - t1)[3] # Time in seconds
> m1  = sum(gc()[,6]) - m1    # Peak excess memory use during the test
> # Fast brute force full correlation:
> m2  = sum(gc()[,2])
> t2  = proc.time()
> mu = colMeans(A)
> s  = sqrt(apply(A, 2, crossprod) - nrow(A) * mu ^ 2)
> cs = scale(A, center=mu, scale=s)  # scale and center the data
> cx = crossprod(cs)                 # full correlation matrix
> cx  = cx * upper.tri(cx)    # ignore symmetry and diagonal
> pairs = which(cx >= threshold, arr.ind=TRUE)
> t2  = (proc.time() - t2)[3]
> m2  = sum(gc()[,6]) - m2
\end{Sinput}
\end{Schunk}
\end{small}
Brute-force computation outperforms {\tt tcor()} in this small first example in
speed but consumes substantially more memory.  Each algorithm identifies the
same set of 125 gene expression pairs that exceed the specified correlation
threshold of 0.95, with comparative timing and memory use shown in the
following table.

% latex table generated in R 3.2.3 by xtable 1.8-0 package
% Sun Dec 20 09:52:31 2015
\begin{table}[ht]
\centering
{\small
\begin{tabular}{lrr}
  \hline
Method & Wall clock time (s) & Peak excess memory use (MB) \\ 
  \hline
tcor & 3.19 & 59.60 \\ 
  brute force & 1.74 & 1075.10 \\ 
   \hline
\end{tabular}
}
\caption{Gene expression results, threshold=0.95} 
\label{EisenYeast}
\end{table}

@@

% latex table generated in R 3.2.3 by xtable 1.8-0 package
% Sun Dec 20 09:52:31 2015
\begin{table}[ht]
\centering
{\small
\begin{tabular}{lrrl}
  \hline
Method & Wall clock time (s) & Memory use (MB) &   \\ 
  \hline
tcor (4 core PC) & 10309.00 & 389.10 &  \\ 
  tcor (64 core cluster) & 482.00 & 994.30 & (local) \\ 
   &  & 42282.40 & (remote) \\ 
   \hline
\end{tabular}
}
\caption{TCGA ACC Methylation450 (80 rows x 394,014 columns, threshold=0.99). Linux cluster results include master R process memory use (local) plus total peak memory used by all 64 worker R processes (remote).} 
\label{TCGA3}
\end{table}From a set of almost 160 billion pairs of vectors, Algorithm \ref{prune} finds 913,601
pairs with correlation values of 0.99 or more on a modest Linux
cluster in about 8 minutes. The same algorithm works, more slowly, in
the constrained memory setting of a desktop PC without modification.
